The Evolution of Phenotypically Invariant Genetic Networks

### 2017-09-14 kalman_walk_paper.tex
added an informal description of the kalman decomposition to the main text. Edited metabolic example and recombination section.
deleted parameter estimate section. edited some of the quantitative genetics stuff -- I think it needs a bit of work though. There are a lot of different versions of the same argument floating around, whether we are using gamma, alpha, c, xi etc.

depending on $u_{2}$ and $u_{1}$ the organism will synthesize an appropriate $\phi_{1}$ and $\phi_{2}$. 

@@

  $\Sys_{\min}(U) = (U \left[ \begin{smallmatrix} 0 & -1 \\ 0 & 0 \end{smallmatrix} \right] U^{-1}, U \left[\begin{smallmatrix} 1 & 0 \\ 0 & 1 \end{smallmatrix}\right], \left[\begin{smallmatrix} 1 & 0 \\ 0 & 1 \end{smallmatrix}\right]U^{-1})$. 

@@

  $\Sys(\small X_{1-3},V) \! =  \! \left( \! \begin{array}{ll}
     V \! \left[\begin{smallmatrix} X_{1} & X_{2} \\ 0 & A \end{smallmatrix}\right] \! V^{-1}, V \! \left[\begin{smallmatrix} X_{3} \\ B \end{smallmatrix}\right], \left[\begin{smallmatrix} 0 & C \end{smallmatrix}\right] \! V^{-1} \end{array} \! \right)

@@

  ($V$ can be any $4$-dimensional and invertible matrix, $X_{1-3} \in \R^{2 \time 2}$, and $(A,B,C) \in \Sys_{\min}$).

@@

    For instance, the diploid $\{\Sys, \Sys' \}$ has system architure $\Sys_{\text{diploid}} \! = \! \left(\Sys \! + \! \Sys'\right)\!/2$, the average of its two haploid systems. During meiosis a diploid recombines its two genomes to form a new one (a gamete): $\left(\Sys \! \times \! \Sys' \right) \rightarrow \Sys''$; by randomly swapping matrix rows between two systems (Appendix \ref{apx:recombination}). Then two gametes join to form an offspring (\emph{e.g.} $\{\Sys'',\Sys\} $).

@@

\begin{figure}[H]
  \centering
  \begin{tabular}{ccc}
    \includegraphics[width=0.25\textwidth, height=0.125\paperheight]{F1_comparison} & 
    \includegraphics[width=0.25\textwidth, height=0.125\paperheight]{F2s_comparison2} &
    \includegraphics[width=0.25\textwidth, height=0.125\paperheight]{figures/quartic_quadratic2}
  \end{tabular}
  \caption{
    $F_1$ (left) and $F_2$ (right) hybrids crossing $\Sys(2)$ with $\Sys(2.01)$ (top), $\Sys(2.1)$ (middle), and $\Sys(2.5)$ (bottom).
    $F_2$ hybrids display more phenotypic divergence than $F_1$s, on average. Further, some $F_2$s completely fail to oscillate, as seen in an $\Sys(2.5)$ $F_2$ (light blue).
    \plri{make axis labels bigger}
  }
\end{figure}
\end{example}
%
%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Systems drift and the accumulation of incompatibilities}
Thus far we have shown that many distinct molecular mechanisms can realize identical phenotypes
and that these mechanisms may fail to produce viable hybrids.
Here we discuss the rate at which evolution shifts molecular mechanisms
and ask whether this process is fast enough to be a significant driver of speciation.
To do so, we explore a general quantitative genetic model in which a population drifts stochastically
near a set of equivalent and optimal systems. We work with a population of effective size $N_e$.

@@

\paragraph{Offspring.}
\js{Not sure if this section is redundant, since reproduction is described above for the LTI model.}
Individuals are diploid; we assume that each haploid genome determines a set of coefficients,
and the individual's coefficients are the average of her two haploid values (no dominance).
This implies that the diploid population variance, $\sigma^2$, is one-half the haploid variance.
Each new gamete is produced from the parent's two haploid copies;
for simplicity we assume that the gamete inherits a random choice of one of the two parental copies,
and so $\sigma$ remains constant, up to a $1/N_e$ term.
A more general model including segregation variance \citep{barton_infinitesimal}
would result in the same qualitative conclusions.
\plri{but put this in the appendix?}

\paragraph{System drift}
% plr how fast does it drift
If the variation within a population of some coefficient
has standard deviation $\sigma$,
then since subsequent generations resample from this diversity,
the population mean coefficient will move a random distance of size $\sigma/\sqrt{N_e}$ per generation, 
\js{since it moves at rate $\sigma \sqrt{T/N_e}$ isn't saying  $\sigma/\sqrt{N_e}$ \emph{per generation} not that clear?}
simply because this is the the standard deviation of the mean of a random sample \citep{lande_drift}.
% (This could be taken as a definition of $N_e$.)
Selection will tend to restrain this motion,
but mean movement along the optimal set $\optx$ is unconstrained.
% (although perhaps complicated by recombination load \citep{recomb_load})
The amount of variance in particular directions in coefficient space 
depend on constraints imposed by selection and 
the covariance between genetic variation between different coefficients (the $G$ matrix \citep{G_matrix}).
% covariance which may arise due to functional constraints and/or statistical linkage.
% There may well be functional constraints -- but these are not sufficiently well-known to say anything general about.
For instance,
if the variation is due to \textit{cis}-regulatory variants,
the genetic basis of each \emph{row} of $A$ likely lies within a few kilobases of tightly linked sequence,
across which a population may carry only a few common haplotypes.
However, covariance due to transiently assembled haplotypes is not expected to be stable over long periods of time --
a common \textit{cis}-regulatory haplotype of transcription factor $k$ with particularly strong binding to both $i$ and $j$
(leading to positive covariance between $A_{ik}$ and $A_{jk}$)
is no more likely to appear than one with strong binding to $i$ but particularly weak binding to $j$ (negative covariance).
(Such transient covariances may well increase the variance of the per-generation change in network mean, however \citep{barton_linkage}.)

@@

To obtain a general quantitative picture, we need to know 
$\sigma_N$ and $\sigma_S$, the standard deviations of coefficient variation along and perpendicular to $\optx$ respectively,
and $\gamma$, the scale on which phenotype changes moving away from $\optx$.
Concretely, $\gamma$ is the inverse of the derivative of $\dph(\Phi(x+u z), \Phi(x))$ 
with respect to $u$ for $x \in \optx$ and $z$ perpendicular to the tangent space at $x$.
With these parameters, a typical individual will have a fitness of around $\exp(-(\sigma_S/\gamma)^2)$.

\paragraph{Hybridization}
\plr{I put F1s and F2s here, and $\Incompat$, but with less explanation than below. Merge somehow.} 
By the arguments above, the means of two allopatric populations separated for $T$ generations
will be a distance of order $2\sigma_N \sqrt{T/N_e}$ apart along $\optx$.
A population of $F_1$ hybrids has one haploid genome from each,
whose coefficients are averaged,
and so will have mean system coefficients at the midpoint between their means,
and variance equal to $\sigma$. 
\js{I think you mean $\sigma^{2}$?}
Each $F_2$ hybrid will be homozygous for one parental allele on averge at half of the loci in the genome,
so the distribution of $F_2$s will have mean at the average of the two populations,
as before,
but variance equal to $\sigma^2 + z^2/2$, where $z$ is the distance between the parental populations. 
\js{Can you clarify why $z^{2}/2$? Is this because it's averaging two distributions?}
These are depicted in figure \ref{fig:conceptual_fig}.

@@

\paragraph{Speciation rates under neutrality}
%
\plr{separate out here (or above) the difference in genotype of hybrids from the optimum set, the difference in phenotype, and the difference in fitness.}
%In general terms, if fitness $\fit$ is some function $g$ of phenotypic divergence from an optimal scaled by the strength of selection $\beta$,
%\begin{align*}
%  \fit = g \left( \frac{\phi - \phi_{0}}{\beta} \right)
%\end{align*}
%\plr{sentences don't begin with "And"}
%And a population $P$ is normally distributed around a mean gene network architecture $A(\tau)$ and $\sigma^{2}$ is the intra-population regulatory variance,
%\plr{what is $T$?}
%\begin{align*}
%  P_{T} \sim \mathcal{N} \left(A(\tau), T \frac{\sigma^{2}}{N_{e}}\right)
%\end{align*}
%the population mean architecture will move at rate
%\plr{what is $\Delta \tau$? Also, if we reframe in terms of $V$ not $\tau$ above will need to reframe this.}
%\begin{align*}
%  \Delta \tau = \sigma \sqrt{\frac{T}{N_{e}}}
%\end{align*}
%where $T$ is time measured in generations. 
%USING GENERAL MATH FROM MULTIVARIATE TAYLOR SERIES.... we know that the phenotype of a first generation $F_{1}$ hybrid diverges from that of its parents (assuming both parents are at an optimum and phenotypically equivalent) quartically as a function of $\Delta \tau$. 
%Thus the squared difference of it's phenotype will be,
%\begin{align*}
%  \Delta \Phi^{2}_{1} = c_{1}^{2} \sigma^{4} \left( \frac{T}{N_{e}} \right)^{2}
%\end{align*}
%and $F_{1}$ fitness will decline linearly with respect to $\frac{T}{N_{e}}$. ARGUMENT FOR WHY $\sigma^{2} c_{1} \sim \beta$ HERE.
%   \begin{align*}
%     \Delta \fit_{1} = \left( \frac{\Delta \Phi_{1}}{\beta} = \frac{c_{1} \sigma^{2} \frac{T}{N_{e}}}{\beta} \approx \frac{T}{N_{e}} \right)
%\end{align*}
% It follows FROM THE SAME MATH ABOUT THE MULTIVAR TAYLOR that the phenotypes of second generation $F_{2}$ hybrid crosses diverges quadratically as a function of $\Delta \tau$, 
%\begin{align*}
%  \Delta \Phi^{2}_{2} &= c_{2}^{2} \left( \sigma \sqrt{\frac{T}{N_{e}}} \right)^{2} = c_{2}^{2} \frac{T}{N_{e}} \\
%  \Delta \fit_{2} &= \left( \frac{\Delta \Phi_{2}}{\beta} = \frac{c_{2} \sigma \sqrt{\frac{T}{N_{e}}}}{\beta} \approx \sqrt{\frac{T}{N_{e}}} \right)
%\end{align*}
%and that $F_{2}$ fitness drops rapidly -- at approximately rate $\sqrt{\frac{T}{N_{e}}}$. ARGUMENT FOR WHY $c_{2} \sigma \sim \beta$ HERE. 
%
For cases where system structures and $\gamma$ can be inferred,
specific speciation rates can be estimated.
For other situations we offer a back of the envelope calculation.

@@

%If the strength of selection is only a few multiples $\xi_1$ of the standing phenotypic diversity within a population $\sigma_S$, such that $\beta = \xi_1 \sigma_S$, and recombination load constrains the standing phenotype-neutral diversity $\sigma_N$ to be only a few multiples $\xi_2$ of $\sigma_S$, then $F_2$ fitness drop will be,
%  \begin{align}
%    \exp\left(- \xi^{2} \frac{T}{N_e} \right)
%  \end{align}
%  where $\xi := 2\xi_2/\xi_1$.
  This suggests that, as an upper bound, if a population contains as much phenotypic diversity as allowed by selection, reproductively isolated populations can speciate rapidly -- possibly on timescales shorter than $N_{e}$ generations.
%%%%%%%%%%%%%%%%%%%%%%%%
\subsection*{Parameter estimates for real systems}
\js{I think we should move this section?}
To translate these results into real predictions, we need to know
the strength of stabilizing selection on the phenotype,
and the amount (and structure) of heritable variation in the genotype.
These are known at best only roughly \citep{felsenstein1988phylogenies},
so we aim for order-of-magnitude estimates.

%% amount of heritable variation
We quantify (roughly) the amount of heritable variation
by $\sigma^2$, the genetic variance present in a population in a typical entry of $A$.
The coefficient $A_{ij}$ measures how much the rate of net production of $i$ changes
per change in concentration of $j$.
It is generally thought that regulatory sequence change contributes much more to inter- and intraspecific variation
than does coding sequence change affecting molecular structure \citep{schmidt2010vertebrate}.
In the context of transcription factor networks this may be affected 
not only by the binding strength of molecule $j$ to the promoter region of gene $i$
but also the effects of other transcription factors (e.g., cooperativity)
and local chromatin accessibility \citep{stefflova2013cooperativity}.
For this reason, 
the mutational target size for variation in $A_{ij}$ may be much larger than the dozens of base pairs
typically implicated in the handful of binding sites for transcription factor $j$ of a typical promoter region,
and single variants may affect many entries of $A$ simultaneously.
% (Recall that although these are best modeled through nonlinear terms,
% by linearizing we essentially consider first-order effects.)
On the other hand, a diverse set of buffering mechanisms are thought to contribute to phenotypic stability
in the presence of substantial molecular noise \citep{canalization,buffering},
suggesting that substantial variation in the micro-scale dynamics we consider here
may be necessary to produce relevant phenotypic effects downstream.
\plr{replace "micro-scale" with sthg else or discuss earlier}

%% empirical data on sigma
The amount and structure of this standing variation is established over long time scales
by many factors, including
mutation-selection balance, 
shifts in the phenotypic optimum,
and/or spatial variation in the optimum \citep{hansen1996translating}.
Quantitative genetics models of mutation-selection balance 
predict precise levels and structure of standing variation \citep{kimura_mutsel,lande_mutsel,lande1981models},
but it is unclear how well these predictions match reality \citep{johnson_barton}
and how much they are expected to change over time \citep{arnold_changing_G}.
However, empirical work allows us to estimate at least the rough magnitude of variation.
Differences in $A_{ij}$ due to a sequence change are hard to measure,
but variation in both transcription factor binding site occupancy
and expression levels (e.g., cis-eQTL) have been measured in various systems.
However, variation in binding site occupancy may overestimate variation in $A$,
since it does not capture buffering effects (if for instance only one site of many needs be occupied for transcription to begin),
and variation in expression levels measures changes in steady-state concentration (our $\kappa_i$) rather than the \emph{rate} of change.
Nonetheless,
\citet{kasowski2010variation} found differential occupancy in 7.5\%
of binding sites of a transcription factor (p65) between human individuals.
\citet{verlaan2009targeted} showed that cis-regulatory variation
accounts for around 2--6\% of expression variation in human blood-derived primary cells,
while \citep{lappalainen2013transcriptome} found that human population variation 
explained about 3\% of expression variation.
\plr{Get some data from at least one other species in here!}
Taken together, this suggests that variation in the entries of $A$
may be on the order of 1\% between individuals of a population --
doubtless varying substantially between species and between genes.

%% argument for strength of selection
It seems certain that selection in most species is not so strong that intra-population variation
is strongly deleterious,
so that if $\beta$ is the typical scale on which selection acts,
then $\beta > \sigma$.  
\plr{do we call this $\beta$ or $u$?} \js{$\beta$, $u$ is input.}
However,
a range of studies \plr{find them} have found evidence for weak stabilizing selection
on regulatory SNPs and cis-eQTL.
For instance, 
\citet{maria_and_sergey} \plr{(others?)} found evidence that large-effect regulatory mutations are weakly selected against in Drosophila.
This suggests that 
the strength of selection on phenotype is sufficient to weakly constrain regulatory variation,
so that perhaps $\sigma$ and $\beta$ are relatively close.
This is as would be expected if available variation is held in check by mutation--selection balance
rather than genetic drift.
A conservative estimate would be that $\beta = 5 \sigma$;
taking $\sigma=.01$ as above, this suggests that changes in phenotype of 5\% are sufficient
to effect a noticeable drop in fitness.
\plr{BUT $\sigma$ IS VARIATION IN $A$ NOT PHENOTYPE}

We have guessed that within a population, 
the entries of $A$ vary by around 1\%, 
at least for networks whose function is strongly constrained.

%%%%%%%%%%%%%%%

>>>>

depending on $u_{1}$ and $u_{2}$ the organism will synthesize an appropriate $\phi_{1}$ and $\phi_{2}$. 

@@

  $\Sys_{\min}(U) = (U \left[ \begin{smallmatrix} 0 & -1 \\ 0 & 0 \end{smallmatrix} \right] U^{-1}, U 
    %\left[\begin{smallmatrix} 1 & 0 \\ 0 & 1 \end{smallmatrix}\right]
      ,
     % \left[\begin{smallmatrix} 1 & 0 \\ 0 & 1 \end{smallmatrix}\right]
        U^{-1})$. 

@@

  $\Sys(\small D_{1-3},V) \! =  \! \left( \! \begin{array}{ll}
     V \! \left[\begin{smallmatrix} D_{1} & D_{2} \\ 0 & A \end{smallmatrix}\right] \! V^{-1}, V \! \left[\begin{smallmatrix} D_{3} \\ B \end{smallmatrix}\right], \left[\begin{smallmatrix} 0 & C \end{smallmatrix}\right] \! V^{-1} \end{array} \! \right)

@@

  ($V$ can be any $4$-dimensional and invertible matrix, $D_{1-3} \in \R^{2 \time 2}$, and $(A,B,C) \in \Sys_{\min}$).

@@

    For instance, the diploid $\{\Sys, \Sys' \}$ has system architecture $\Sys_{\text{diploid}} \! = \! \left(\Sys \! + \! \Sys'\right)\!/2$, the average of its two haploid systems. During meiosis a diploid recombines its two genomes to form a new one (a gamete): $\left(\Sys \! \times \! \Sys' \right) \rightarrow \Sys''$; by randomly swapping matrix rows between two systems (Appendix \ref{apx:recombination}). Then two gametes join to form an offspring (\emph{e.g.} $\{\Sys'',\Sys\} $).

@@

\begin{figure}[H]
  \centering
  \begin{tabular}{ccc}
    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F1_comparison} & 
    \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{F2s_comparison2} \\
    \includegraphics[width=0.25\textwidth, height=0.125\paperheight]{figures/f1_quartic2} &
    \includegraphics[width=0.25\textwidth, height=0.125\paperheight]{figures/f2_quad2} 
  \end{tabular}
  \caption{
    $F_1$ (left) and $F_2$ (right) hybrids crossing $\Sys(2)$ with $\Sys(2.01)$ (top), $\Sys(2.1)$ (middle), and $\Sys(2.5)$ (bottom).
    $F_2$ hybrids display more phenotypic divergence than $F_1$s, on average. Further, some $F_2$s completely fail to oscillate, as seen in an $\Sys(2.5)$ $F_2$ (light blue).
    \plri{make axis labels bigger}
  }
\end{figure}
\end{example}
%
%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Systems drift and the accumulation of incompatibilities}
Thus far we have shown that many distinct molecular mechanisms can realize identical phenotypes
and that these mechanisms may fail to produce viable hybrids.
Here we discuss the rate at which evolution shifts molecular mechanisms
and ask whether this process is fast enough to be a significant driver of speciation.
To do so, we explore a general quantitative genetic model in which a population drifts stochastically
near a set of equivalent and optimal systems. We work with a population of effective size $N_e$.

@@

%\paragraph{Offspring.}
%\js{Not sure if this section is redundant, since reproduction is described above for the LTI model.}
%Individuals are diploid; we assume that each haploid genome determines a set of coefficients,
%and the individual's coefficients are the average of her two haploid values (no dominance).
%This implies that the diploid population variance, $\sigma^2$, is one-half the haploid variance.
%Each new gamete is produced from the parent's two haploid copies;
%for simplicity we assume that the gamete inherits a random choice of one of the two parental copies,
%and so $\sigma$ remains constant, up to a $1/N_e$ term.
%A more general model including segregation variance \citep{barton_infinitesimal}
%would result in the same qualitative conclusions.
%\plri{but put this in the appendix?}

\paragraph{System drift}
% plr how fast does it drift
If the regulatory coefficient population variation
has standard deviation $\sigma$,
since subsequent generations resample from this diversity,
the population mean coefficient will move a random distance of size $\sigma/\sqrt{N_e}$ per generation,
\js{since it moves at rate $\sigma \sqrt{T/N_e}$ isn't saying  $\sigma/\sqrt{N_e}$ \emph{per generation} not that clear?}
simply because this is the the standard deviation of the mean of a random sample \citep{lande_drift}.
% (This could be taken as a definition of $N_e$.)
Selection will tend to restrain this motion,
but mean movement along the optimal set $\optx$ is unconstrained.
% (although perhaps complicated by recombination load \citep{recomb_load})
The amount of variance in particular directions in coefficient space 
depend on constraints imposed by selection and 
the covariance between genetic variation between different coefficients (the $G$ matrix \citep{G_matrix}).
% covariance which may arise due to functional constraints and/or statistical linkage.
% There may well be functional constraints -- but these are not sufficiently well-known to say anything general about.
%For instance,
%if the variation is due to \textit{cis}-regulatory variants,
%the genetic basis of each \emph{row} of $A$ likely lies within a few kilobases of tightly linked sequence,
%across which a population may carry only a few common haplotypes.
%However, covariance due to transiently assembled haplotypes is not expected to be stable over long periods of time --
%a common \textit{cis}-regulatory haplotype of transcription factor $k$ with particularly strong binding to both $i$ and $j$
%(leading to positive covariance between $A_{ik}$ and $A_{jk}$)
%is no more likely to appear than one with strong binding to $i$ but particularly weak binding to $j$ (negative covariance).
%(Such transient covariances may well increase the variance of the per-generation change in network mean, however \citep{barton_linkage}.)

@@

%To obtain a general quantitative picture, we need to know 
We denote
$\sigma_N$ and $\sigma_S$, the standard deviations of coefficient variation along and perpendicular to $\optx$ respectively,
and $\gamma$, the scale on which phenotype changes moving away from $\optx$.
Concretely, $\gamma$ is the inverse of the derivative of $\dph(\Phi(x+u z), \Phi(x))$ 
with respect to $u$ for $x \in \optx$ and $z$ perpendicular to the tangent space at $x$.
With these parameters, a typical individual will have a fitness of around $\exp(-(\sigma_S/\gamma)^2)$.

\paragraph{Hybridization}
\plr{I put F1s and F2s here, and $\Incompat$, but with less explanation than below. Merge somehow.} 
The means of two allopatric populations separated for $T$ generations
will be a distance of order $2\sigma_N \sqrt{T/N_e}$ apart along $\optx$.
A population of $F_1$ hybrids has one haploid genome from each,
whose coefficients are averaged,
and so will have mean system coefficients at the midpoint between their means,
and variance equal to $\sigma^2$. 
\js{I think you mean $\sigma^{2}$? I changed it.}
Each $F_2$ hybrid will be homozygous for one parental allele on averge at half of the loci in the genome,
so the distribution of $F_2$s will have mean at the average of the two populations,
as before,
but variance equal to $\sigma^2 + z^2/2$, where $z$ is the distance between the parental populations. 
\js{Can you clarify why $z^{2}/2$? Is this because it's averaging two distributions?}
These are depicted in figure \ref{fig:conceptual_fig}.

@@

If we prefer not to make any assumptions about the shape of the fitness function, we can more conservatively state that hybrid phenotypes diverge from their optimal parents at rates $d(\Phi_{F_1}, \Phi_0) = \alpha_1 \sigma_N^2 T/N_e$ and $d(\Phi_{F_2}, \Phi_0) = \alpha_2 \sigma_N \sqrt{T/N_e}$, where $\alpha_1$ and $\alpha_2$ are constants. Whether or not hybrid phenotypic divergence is associated with a high fitness cost, $F_2$ phenotypes should diverge much faster than $F_1$s, with a rate proportional to the square root of $T/N_e$, versus linearly.
\js{I think $\alpha_1$ is your $c_1$ and $\alpha_2$ is a function of $\gamma^{-1}$?}

%\subsection*{Parameter estimates for real systems}
%\js{I think we should move this section?}
%To translate these results into real predictions, we need to know
%the strength of stabilizing selection on the phenotype,
%and the amount (and structure) of heritable variation in the genotype.
%These are known at best only roughly \citep{felsenstein1988phylogenies},
%so we aim for order-of-magnitude estimates.

%% amount of heritable variation
%We quantify (roughly) the amount of heritable variation
%by $\sigma^2$, the genetic variance present in a population in a typical entry of $\Sys$.
%The coefficient $A_{ij}$ measures how much the rate of net production of $i$ changes
%per change in concentration of $j$.
%It is generally thought that regulatory sequence change contributes much more to inter- and intraspecific variation
%than does coding sequence change affecting molecular structure \citep{schmidt2010vertebrate}.
%In the context of transcription factor networks this may be affected 
%not only by the binding strength of molecule $j$ to the promoter region of gene $i$
%but also the effects of other transcription factors (e.g., cooperativity)
%and local chromatin accessibility \citep{stefflova2013cooperativity}.
%For this reason, 
%the mutational target size for variation in $A_{ij}$ may be much larger than the dozens of base pairs
%typically implicated in the handful of binding sites for transcription factor $j$ of a typical promoter region,
%and single variants may affect many entries of $\Sys$ simultaneously.
%% (Recall that although these are best modeled through nonlinear terms,
%% by linearizing we essentially consider first-order effects.)
%On the other hand, a diverse set of buffering mechanisms are thought to contribute to phenotypic stability
%in the presence of substantial molecular noise \citep{canalization,buffering},
%suggesting that substantial variation in the micro-scale dynamics we consider here
%may be necessary to produce relevant phenotypic effects downstream.
%\plr{replace "micro-scale" with sthg else or discuss earlier}

%% empirical data on sigma
%The amount and structure of this standing variation is established over long time scales
%by many factors, including
%mutation-selection balance, 
%shifts in the phenotypic optimum,
%and/or spatial variation in the optimum \citep{hansen1996translating}.
%Quantitative genetics models of mutation-selection balance 
%predict precise levels and structure of standing variation \citep{kimura_mutsel,lande_mutsel,lande1981models},
%but it is unclear how well these predictions match reality \citep{johnson_barton}
%and how much they are expected to change over time \citep{arnold_changing_G}.
%However, empirical work allows us to estimate at least the rough magnitude of variation.
%Differences in $A_{ij}$ due to a sequence change are hard to measure,
%but variation in both transcription factor binding site occupancy
%and expression levels (e.g., cis-eQTL) have been measured in various systems.
%However, variation in binding site occupancy may overestimate variation in $\Sys$,
%since it does not capture buffering effects (if for instance only one site of many needs be occupied for transcription to begin),
%and variation in expression levels measures changes in steady-state concentration (our $\kappa_i$) rather than the \emph{rate} of change.
%%Nonetheless,
%%\citet{kasowski2010variation} found differential occupancy in 7.5\%
%%of binding sites of a transcription factor (p65) between human individuals.
%%\citet{verlaan2009targeted} showed that cis-regulatory variation
%%accounts for around 2--6\% of expression variation in human blood-derived primary cells,
%%while \citep{lappalainen2013transcriptome} found that human population variation 
%%explained about 3\% of expression variation.
%%\plr{Get some data from at least one other species in here!}
%%Taken together, this suggests that variation in the entries of $\Sys$
%%may be on the order of 1\% between individuals of a population --
%%doubtless varying substantially between species and between genes.
%%
%Given the empirical data on transcription factor binding and gene expression
%intrapopulation variation \citet{kasowski2010variation, verlaan2009targetedm lappalainen2013transcriptome}, we conservatively estimate that variation in the regulatory coefficients in $\Sys$ may be on the order of 1\% between individuals of a population -- 
%doubtless varying substantially between species and between genes.
%
%%% argument for strength of selection
%%It seems certain that selection in most species is not so strong that intra-population variation
%%is strongly deleterious,
%%so that if $\beta$ is the typical scale on which selection acts,
%%then $\beta > \sigma$.  
%%However,
%%a range of studies \plr{find them} have found evidence for weak stabilizing selection
%%on regulatory SNPs and cis-eQTL.
%For instance, 
%\citet{maria_and_sergey} \plr{(others?)} found evidence that large-effect regulatory mutations are weakly selected against in Drosophila.
%This suggests that 
%the strength of selection on phenotype is sufficient to weakly constrain regulatory variation,
%so that perhaps $\sigma$ and $\beta$ are relatively close.
%This is as would be expected if available variation is held in check by mutation--selection balance
%rather than genetic drift.
%A conservative estimate would be that $\beta = 5 \sigma$;
%taking $\sigma=.01$ as above, this suggests that changes in phenotype of 5\% are sufficient
%to effect a noticeable drop in fitness.
%\plr{BUT $\sigma$ IS VARIATION IN $A$ NOT PHENOTYPE}
%
%We have guessed that within a population, 
%the entries of $\Sys$ vary by around 1\%, 
%at least for networks whose function is strongly constrained.
%
%
\paragraph{Speciation rates under neutrality}
For cases where system structures and $\gamma$ can be inferred,
specific speciation rates can be estimated.
For other situations we offer a back of the envelope calculation.

@@

%If the strength of selection is only a few multiples $\xi_1$ of the standing phenotypic diversity within a population $\sigma_S$, such that $\beta = \xi_1 \sigma_S$, and recombination load constrains the standing phenotype-neutral diversity $\sigma_N$ to be only a few multiples $\xi_2$ of $\sigma_S$, then $F_2$ fitness drop will be,
%  \begin{align}
%    \exp\left(- \xi^{2} \frac{T}{N_e} \right)
%  \end{align}
%  where $\xi := 2\xi_2/\xi_1$.
  This suggests that, as an upper bound, if a population contains as much phenotypic diversity as allowed by selection, reproductively isolated populations can speciate rapidly -- possibly on timescales shorter than $N_{e}$ generations (It is likely that, on average, $c \leq 2/3$).
%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%
